MULTI EPOCH SPATIALLY RESOLVED RADIO OBSERVATIONS OF BETELGEUSE'S WIND ACCELERATION REGION

### 2014-07-11 draft1/ms.tex
Jul 11 Friday eve from Onsala

		 2004 Oct 21,30  & 0.7 & $39\times 26$ & 0.37	& $99\pm 3$ & $92 \pm 20 $& $92\pm 20$ &$28.68\pm 0.53$ & $2940\pm 170$\\
		 				 & 1.3 & $80\times 42$ & 0.09   & $127\pm 2$& $ 114\pm 4$ & $85\pm 5$& $13.83\pm 0.10$& $3140\pm 80$\\
						 & 2.0 & $121\times 91$ & 0.08  & $158\pm 6$& $ 152\pm 17$ & $132\pm 59$ & $7.23\pm 0.15$& $2270\pm 130$ \\
						 & 3.6 & $208\times 126$ & 0.02	& $215\pm 7$& $188\pm 15$ & $0\pm 10$& $3.34\pm 0.03$& $2110\pm 110$\\
						 & 6.1 & $377\times 264$ & 0.02	& $314\pm 30$& $204\pm 54$ & $0\pm 20$ & $1.55\pm 0.04$& $1140\pm 160$\\

@@

\rule{-2.6pt}{2.5ex}  2003 Aug 10,12 & 0.7 		& $40\times 27$ & 0.46	& $103\pm 4$& $92\pm 10$ & $104\pm 16$& $28.05\pm 0.84$& $2760\pm 230$ \\
									 & 1.3		& $80\times 42$ & 0.17	& $122\pm 5$& $115\pm 12$& $0\pm 5$& $11.20\pm 0.24$& $2490\pm 150$\\
									 & 2.0		& $119\times 96$& 0.10	& $132\pm 10$& $ 116\pm 22$& $11\pm 27$& $5.88\pm 0.17$&$3040\pm 360$\\
									 & 3.6		& $ 204\times 139$&	0.03& $193\pm 7$& $ 141\pm 17$ & $152\pm 7$& $2.80\pm 0.04$&  $2610\pm 170$\\
									 & 6.1 		& $378\times 297$&	0.03& $247\pm 47$& $180\pm 103$ & $169\pm 28$& $1.22\pm 0.04$&$2040\pm 680$ \\

@@

\rule{-2.6pt}{2.5ex}  2002 Apr 12,13  & 1.3 		& $91\times 59$ & 0.18	&$134\pm 9$ & $102\pm 16$& $36\pm 10$& $8.96\pm 0.24$& $2170\pm 250$\\
							& 2.0		&$131\times 98$ & 0.39	& $166\pm 16$& $105\pm 27$&$41\pm 11$ & $5.32\pm 0.24$& $2420\pm 450$ \\
							& 3.6		& $224\times 155$&0.03	& $234\pm 9$& $171\pm 18$& $41\pm7$& $2.66\pm 0.04$& $1690\pm 110$\\

@@

\rule{-2.6pt}{2.5ex}  2002 Feb 17,18 & 1.3 		& $83\times 48$&0.14	&$120\pm 4$ & $109 \pm 8$& $31\pm 13$&$10.87\pm 0.17$ & $2750\pm 140$\\
									& 2.0		& $128\times 90$&0.11	&$140\pm 13$& $136\pm 33$& $0\pm 28$&$5.38\pm 0.22$ & $2150\pm 300$\\
									& 3.6		& $200\times 135$&	0.03&$217\pm 9$ & $182\pm 21 $& $ 0\pm 5$ & $2.85\pm 0.04$& $1830\pm 110$\\

@@

\rule{-2.6pt}{2.5ex}  2001 Jan 02  & 1.3 		& $78\times 42$&0.08	&$124\pm 2$ & $114\pm 4$& $40\pm 8$& $12.58\pm0.08$& $2920\pm 70$ \\
 		 2000 Dec 23 & 0.7		& $44\times 20$& 0.18	&$98\pm 2$ &$90\pm 4$ &$0\pm 8$ & $29.02\pm 0.30$& $3070\pm 100$

@@

\tablenotetext{}{Notes.- The restoring beam sizes and image rms noise values are taken from the uniformly weighted radio maps that include the PT antenna baselines. The position angles (measured in degrees east of north) of the restoring beams are not quoted here, but are all between $30^{\circ}$ and $70^{\circ}$. The major and minor axis of the stellar radio disk $\theta _{\mathrm{maj}}$ and $\theta _{\mathrm{min}}$, the position angle P.A., and the total flux density $F_{\nu}$, are all derived from the best-fit uniform-brightness ($T_b$) elliptical-disk models.}

@@

\subsection{Calibrated Visibilities}
Betelgeuse's position was initially offset by a few restoring beams from the interferometer phase center to avoid any possible spurious artefacts which might accumulate there. One such artefact was noticed in the 2.0\,cm image from April 2002 but our target was sufficiently far away and was not contaminated. We fitted elliptical Gaussian brightness distributions to the source at each wavelength and epoch and measured the position of the peak emission. We then used CASA's \textit{fixvis} task to shift the visibilities such that this peak emission position was placed at phase center. The Python-based task \textit{uvmultifit} \citep{marti_vidal_2014} was then used to fit elliptical disks to the calibrated visibilities, with the major axis ($\theta _{\rm{maj}}$), axis ratio ($\theta _{\rm{min}}/\theta _{\rm{maj}}$, where $\theta _{\rm{min}}$ is the minor axis), position angle (P.A.), and flux density ($F_{\nu}$) left as free parameters. We bounded the axis ratio between 0 and 1, and the position angle between 0 and 180$^{\circ}$.

\section{RESULTS} 
\subsection{Radio Maps}
Radio maps were produced at all wavelengths using uniform weighting to utilize the full capabilities of the PT baselines and achieve maximum spatial resolution. The corresponding restoring beam sizes and image rms noise levels are provided in columns 3 and 4 of Table \ref{tab1}. Calibrations issues with the two 6.1\,cm data sets from 2002 resulted in rms noise values being ~10 times higher than expected. This resulted in the source not being detected in the April 2002 map and the source being located $\sim 300$\,mas away from its expected position in the February 2002 map. We therefore do not include these 6.1\,cm results in Table \ref{tab1} and in any of the analysis.The remainder of the maps were generally of excellent quality with the source clearly resolved at the shorter wavelengths. No deviations from axially symmetry were present in any of the maps between 1.3 and 20.5\,cm.

All of the 0.7\,cm maps showed significant deviations from axial symmetry. However, when the data sets were imaged using the different weighting schemes described in Section \ref{obs2.1}, the morphology was inconsistent. For example, the maps created using standard uniform weighing were morphologically different to the maps created using natural weighting with a restoring beam of size corresponding to that produced with uniform weighting. These inconsistencies are probably due to poor phase stability resulting from rapid tropospheric phase fluctuations of which our cycle times were not short enough to compensate for. To investigate this further, we removed the PT antenna baselines and again imaged the data using the two aforementioned
imaging techniques. We found that two data sets (i.e., from 2000 and 2004) produced images that had consistent morphology when any of the imaging techniques were applied. The observing logs of the other data sets report poorer weather conditions in agreement with our imaging analysis. The resulting images are plotted in Figure \ref{fig1} along with the 0.7\,cm image from \cite{lim_1998} for comparison. In these data sets, the cycle time was adequate to obtain reasonable phase transfer from the calibrator to the source but shorter cycle times were probably required for the other data sets due to poorer weather. Unlike the \cite{lim_1998} map, no asymmetries are present in either of our two maps as is clearly evident in Figure \ref{fig1}. These maps indicate that large scale asymmetries are not always present in Betelgeuse's inner atmosphere.

\begin{figure*}
\mbox{
\includegraphics[trim = 6mm 0mm 0mm 0mm, clip,scale=0.95]{/home/eamon/pietown/paper/figs/fig1.ps}
%\includegraphics[trim = 30mm 90mm 50mm 80mm, clip,scale=0.55]{/home/eamon/pietown/paper/figs/nature_fig.ps}
%\includegraphics[trim = 0mm 50mm 20mm 58mm, clip,scale=0.4,angle=90]{/home/eamon/pietown/paper/figs/vla_q_2000.ps}
%\includegraphics[trim = 0mm 0mm 20mm 58mm, clip,scale=0.40,angle=90]{/home/eamon/pietown/paper/figs/vla_q_2004.ps}
}
\caption{VLA A configuration maps of Betelgeuse at 0.7\,cm over three different epochs. These maps have been created by naturally weighting the visibilities and using a restoring beam corresponding to the size created from uniform weighting. The restoring beam size and shape is located in the bottom left position of each panel while the red circle is the approximate location and size of the optical photosphere. From left to right the beam sizes are x, y, and z$\prime\prime$ and the rms noise values are x, y, and z\,mJy\,beam$^{-1}$. \textit{Left panel}: The \cite{lim_1998} map shows an asymmetry in the east direction. The position of the photosphere is here assumed to be .... \textit{Middle and right panels}: Our 0.7\,cm maps created without any visibilities from the PT baselines. No asymetries deviating from the restoring beam shape are present. The position of the photosphere is here assumed to be located on the peak emission.}
\label{fig1}
\end{figure*}

\subsection{Diameters} 
Importance of working in the visibility domain
Previous VLA observations have only spatially resolved the stellar atmosphere at 0.7\,cm but here we fully resolve the star at all wavelengths between 0.7 and 6.1\,cm. Skinner discussion
Point to Table 1 give diameter variability.

xaxis 'Real Visibility (mJy)' colour pietown baselines

Compare to Harper 2001 paper

\subsection{Flux Densities}
obtained from visibilty fits but also checked against elliptical Gaussian fitted results
plot flux versus frequency (see harper paris presentation)
\subsection{Temperature Profiles}
give more exact lim formula

@@

\subsection{Structure of Wind Acceleration Region} 
Thermal structure (Lim vs vs Ours vs e-MERLIN)\\
see Reid \& Menton 1996 conf proceedings\\
Harper model

>>>>

		 2004 Oct 21,30  & 0.7 & $39\times 26$ & 0.37	& $99\pm 3$ & $92 \pm 3$& $92\pm 20$ &$28.68\pm 0.53$ & $2940\pm 170$\\
		 				 & 1.3 & $80\times 42$ & 0.09   & $127\pm 2$& $ 114\pm 2$ & $85\pm 5$& $13.83\pm 0.10$& $3140\pm 80$\\
						 & 2.0 & $121\times 91$ & 0.08  & $158\pm 6$& $ 152\pm 10$ & $132\pm 59$ & $7.23\pm 0.15$& $2270\pm 130$ \\
						 & 3.6 & $208\times 126$ & 0.02	& $215\pm 7$& $188\pm 7$ & $0\pm 10$& $3.34\pm 0.03$& $2110\pm 110$\\
						 & 6.1 & $377\times 264$ & 0.02	& $314\pm 30$& $203\pm 30$ & $0\pm 20$ & $1.55\pm 0.04$& $1140\pm 160$\\

@@

\rule{-2.6pt}{2.5ex}  2003 Aug 10,12 & 0.7 		& $40\times 27$ & 0.46	& $103\pm 4$& $92\pm 5$ & $104\pm 16$& $28.05\pm 0.84$& $2760\pm 230$ \\
									 & 1.3		& $80\times 42$ & 0.17	& $122\pm 5$& $114\pm 6$& $0\pm 5$& $11.20\pm 0.24$& $2490\pm 150$\\
									 & 2.0		& $119\times 96$& 0.10	& $132\pm 10$& $ 116\pm 9$& $11\pm 27$& $5.88\pm 0.17$&$3040\pm 360$\\
									 & 3.6		& $ 204\times 139$&	0.03& $193\pm 7$& $ 140\pm 11$ & $152\pm 7$& $2.80\pm 0.04$&  $2610\pm 170$\\
									 & 6.1 		& $378\times 297$&	0.03& $247\pm 47$& $179\pm 61$ & $169\pm 28$& $1.22\pm 0.04$&$2040\pm 680$ \\

@@

\rule{-2.6pt}{2.5ex}  2002 Apr 12,13  & 1.3 		& $91\times 59$ & 0.18	&$134\pm 9$ & $102\pm 6$& $36\pm 10$& $8.96\pm 0.24$& $2170\pm 250$\\
							& 2.0		&$131\times 98$ & 0.39	& $166\pm 16$& $104\pm 13$&$41\pm 11$ & $5.32\pm 0.24$& $2420\pm 450$ \\
							& 3.6		& $224\times 155$&0.03	& $234\pm 9$& $171\pm 10$& $41\pm7$& $2.66\pm 0.04$& $1690\pm 110$\\

@@

\rule{-2.6pt}{2.5ex}  2002 Feb 17,18 & 1.3 		& $83\times 48$&0.14	&$120\pm 4$ & $109 \pm 3$& $31\pm 13$&$10.87\pm 0.17$ & $2750\pm 140$\\
									& 2.0		& $128\times 90$&0.11	&$140\pm 13$& $135\pm 16$& $0\pm 28$&$5.38\pm 0.22$ & $2150\pm 300$\\
									& 3.6		& $200\times 135$&	0.03&$217\pm 9$ & $183\pm 10 $& $ 0\pm 5$ & $2.85\pm 0.04$& $1830\pm 110$\\

@@

\rule{-2.6pt}{2.5ex}  2001 Jan 02  & 1.3 		& $78\times 42$&0.08	&$124\pm 2$ & $115\pm 1$& $40\pm 8$& $12.58\pm0.08$& $2920\pm 70$ \\
 		 2000 Dec 23 & 0.7		& $44\times 20$& 0.18	&$98\pm 2$ &$91\pm 1$ &$0\pm 8$ & $29.02\pm 0.30$& $3070\pm 100$

@@

\tablenotetext{}{Notes.- The restoring beam sizes and image rms noise values are taken from the uniformly weighted radio maps that include the PT antenna baselines. The position angles (measured in degrees east of north) of the restoring beams are not quoted here, but are all between $30^{\circ}$ and $70^{\circ}$. The major and minor axis of the stellar radio disk $\theta _{\mathrm{maj}}$ and $\theta _{\mathrm{min}}$, the position angle P.A., and the total flux density $F_{\nu}$, are all derived from the best-fit uniform-brightness ($T_b$) elliptical-disk models. The atmosphere at 20.5\,cm was unresolved so we define the minor axis of the restoring beam as the upper limit to its diameter.}

@@

\subsection{Calibrated Visibilities}\label{obs2.2}
Betelgeuse's position was initially offset by a few restoring beams from the interferometer phase center to avoid any possible spurious artefacts which might accumulate there. One such artefact was noticed in the 2.0\,cm image from April 2002 but our target was sufficiently far away and was not contaminated. We fitted elliptical Gaussian brightness distributions to the source at each wavelength and epoch and measured the position of the peak emission. We then used CASA's \textit{fixvis} task to shift the visibilities such that this peak emission position was placed at phase center. The Python-based task \textit{uvmultifit} \citep{marti_vidal_2014} was then used to fit elliptical disks to the calibrated visibilities, with the major axis ($\theta _{\rm{maj}}$), axis ratio ($\theta _{\rm{min}}/\theta _{\rm{maj}}$, where $\theta _{\rm{min}}$ is the minor axis), position angle (P.A.), and flux density ($F_{\nu}$) left as free parameters. We bounded the axis ratio between 0 and 1, and the position angle between 0 and 180$^{\circ}$.

\section{RESULTS} 
\subsection{Radio Maps}
Radio maps were produced at all wavelengths using uniform weighting to utilize the full capabilities of the PT baselines and achieve maximum spatial resolution. The corresponding restoring beam sizes and image rms noise levels are provided in columns 3 and 4 of Table \ref{tab1}. Calibrations issues with the two 6.1\,cm data sets from 2002 resulted in rms noise values being ~10 times higher than expected. This resulted in the source not being detected in the April 2002 map and the source being located $\sim 300$\,mas away from its expected position in the February 2002 map. We therefore do not include these 6.1\,cm results in Table \ref{tab1} and in any of the analysis.The remainder of the maps were generally of excellent quality with the source clearly resolved at the shorter wavelengths. No significant deviations from $\pi$-rotational symmetry were present in any of the maps between 1.3 and 20.5\,cm and no `radio hotspots' were present.

All of the 0.7\,cm maps showed significant deviations from $\pi$-rotational symmetry. However, when these data sets were imaged using the different weighting schemes described in Section \ref{obs2.1}, the morphology was inconsistent. For example, the maps created using standard uniform weighing were morphologically different to the maps created using natural weighting with a restoring beam of size corresponding to that obtained with uniform weighting. These inconsistencies are probably due to poor phase stability resulting from rapid tropospheric phase fluctuations of which our cycle times were not short enough to compensate for. To investigate this further, we removed the PT antenna baselines and again imaged the data using the two aforementioned imaging techniques. We found that two data sets (i.e., from 2000 and 2004) produced images that had consistent morphology when any of the imaging techniques were applied. The observing logs of the other data sets report poorer weather conditions in agreement with our imaging analysis. The resulting images are plotted in Figure \ref{fig1} along with the 0.7\,cm image from \cite{lim_1998} for comparison. In these data sets, the cycle time was adequate to obtain reasonable phase transfer from the calibrator to the source but shorter cycle times were probably required for the other data sets due to poorer weather. Unlike the \cite{lim_1998} map, no asymmetries are present in either of our two maps as is clearly evident in Figure \ref{fig1}. These maps indicate that large scale asymmetries are not always present in Betelgeuse's inner atmosphere.

\begin{figure*}
\mbox{
\includegraphics[trim = 6mm 0mm 0mm 0mm, clip,scale=0.95]{/home/eamon/pietown/paper/figs/fig1.ps}
%\includegraphics[trim = 30mm 90mm 50mm 80mm, clip,scale=0.55]{/home/eamon/pietown/paper/figs/nature_fig.ps}
%\includegraphics[trim = 0mm 50mm 20mm 58mm, clip,scale=0.4,angle=90]{/home/eamon/pietown/paper/figs/vla_q_2000.ps}
%\includegraphics[trim = 0mm 0mm 20mm 58mm, clip,scale=0.40,angle=90]{/home/eamon/pietown/paper/figs/vla_q_2004.ps}
}
\caption{VLA A configuration maps of Betelgeuse at 0.7\,cm over three different epochs. These maps have been created by naturally weighting the visibilities and applying a restoring beam corresponding to the size obtained from uniform weighting. The restoring beam size and shape is located in the bottom left corner of each panel while the red circle is the approximate location and size of the optical photosphere. The contour levels are plotted at $10\%, 20\%,\dots ,90\%$, and $99\%$ of the peak flux. From left to right the beam sizes are $40^{\prime\prime}\times 40^{\prime\prime}$, $45^{\prime\prime}\times 41^{\prime\prime}$, and $48^{\prime\prime}\times 40^{\prime\prime}$, and the rms noise values are x, 0.18, and 0.34\,mJy\,beam$^{-1}$. \textit{Left panel}: The \cite{lim_1998} map shows an asymmetry in the east direction. The position of the photosphere is here assumed to be .... \textit{Middle and right panels}: Our maps show no significant deviations from the shape of the restoring beam. The position of the photosphere is here assumed to be located on the peak emission.}
\label{fig1}
\end{figure*}

\subsection{Radio Diameters} 
\begin{figure}
\includegraphics[trim = 3mm 0mm 0mm 0mm, clip,height=12cm, width=8.5cm]{/home/eamon/pietown/paper/figs/fig2.ps}
\caption{Real component of the complex visibility plotted against the projected baseline length in wavelengths, $B_{\lambda}$, for the 2004 epoch at all wavelengths between $0.7-6.1$\,cm. The black data point contain only VLA antenna baselines, the red data points contain both VLA and PT antenna baselines, while the green data points contain only PT antenna baselines. The solid blue line represents the theoretical visibilities of a uniform intensity circular disk with an \textit{effective radius} defined as $\phi _{\rm{eff}}=0.5\sqrt{\theta _{\rm{maj}}\theta _{\rm{min}}}$.}
\label{fig2}
\end{figure}
It can be seen in Figure \ref{fig1} that the size of the source at 0.7\,cm is only a few beam diameters across. The source to beam ratio gets progressively smaller at longer wavelengths which makes fitting models to the images unreliable. In this case it is best to obtain estimates of the source size and flux density by fitting models directly to the calibrated complex visibilities. In Figure \ref{fig2} we plot the real component of the complex visibility against the projected baseline length in wavelengths, $B_{\lambda}$, for the 2004 epoch at all wavelengths between $0.7-6.1$\,cm. The data has been azimuthally vector averaged into uniformly sized bins over large portions of the unevenly sampled $u-v$ plane, where $u$ and $v$ are the spatial coordinates of the projected baseline in wavelengths, and $B_{\lambda} = \sqrt{u^2 + v^2}$. It is clear that the source is fully resolved at all wavelengths shown, i.e., we have obtained visibilities close to the first null. The VLA observations of \cite{lim_1998} only fully resolved Betelgeuse's atmosphere at 0.7\,cm at one epoch, but here we fully resolve the star at all wavelengths between 0.7 and 6.1\,cm at all our epochs. The black data points in Figure \ref{fig2} represent the real visibility measurements from the VLA antenna baselines only, the red data points include both VLA and PT antenna baselines, while the green data points include only PT antenna baselines. This highlights the importance of the inclusion of the PT antenna, especially at the longer wavelengths where it provides many baseline pairs close to the first null which helps to constrain the size of the extended atmosphere. The corresponding binned imaginary components of the complex visibility are not shown here but were mainly consistent with zero amplitude at all baselines, indicative of no significant asymmetrical structure.

Uniform intensity elliptical disk models were fitted to the complex visibilities as outlined in Section \ref{obs2.2}. The diameters and P.A.s of the best fits are listed in columns 5, 6, and 7 of Table \ref{tab1}. The source size is found to get progressively larger with increasing wavelength which implies that the radio surfaces are optically thick, in agreement with the findings of \cite{lim_1998}. At 0.7\,cm, these radio surfaces only slightly depart from spherical symmetry having axis ratios $\gtrsim 0.9$, in good agreement with the radio maps. This is also the case at 1.3 and 2.0\,cm except in 2002 April where the axis ratios change to $0.76\pm 0.07$ and $0.63\pm 0.10$ respectively, from near spherical symmetric values two months prior. At 3.5 and 6.1\,cm the radio surfaces display more consistent deviations from spherical symmetry having axis ratios ranging between $0.65\pm 0.10$ and $0.87\pm 0.04$. The intrinsic beam is elongated at all wavelengths due to inclusion of the PT antenna so caution must be exercised when reporting deviations from spherical symmetry.

In Figure \ref{fig2} we also plot the theoretical visibility curves of a uniform intensity circular disk with an \textit{effective radius} defined as $\phi _{\rm{eff}}=0.5\sqrt{\theta _{\rm{maj}}\theta _{\rm{min}}}$. The scatter between these model curves and the azimuthally averaged real visibility is low and indicates that circular disks represent the radio surfaces well. This is found to be the case at all epochs. Bearing in mind that the data is sparsely sampled between 2000 December and 2004 October we can cautiously investigate how the size of the atmosphere changes over this time frame. At 0.7 and 1.3\,cm there is no significant change in $\phi _{\rm{eff}}$ at all epochs with $\Delta \phi _{\rm{eff}} \lesssim 3\%$ about the mean, i.e., there is no significant change in the size of the inner atmosphere. We see more variability at longer wavelengths which sample the more extended atmosphere with $\Delta \phi _{\rm{eff}} \sim 14\%$ about the mean at 2.0 and 3.5\,cm. In fact at these two wavelengths, $\phi _{\rm{eff}}$ increases by $\sim 24\%$ in the 14 months between our 2003 and 2004 observations. We only have two measurements at 6.1\,cm and these again show an increase in $\phi _{\rm{eff}}$ between 2003 and 2004 with $\Delta \phi _{\rm{eff}} =20\%$, although the error are too large be be certain of this increase at 6.1\,cm.

\subsection{Flux Densities}
obtained from visibilty fits but also checked against elliptical Gaussian fitted results
plot flux versus frequency (see harper paris presentation)
$F_{\nu} = 0.185$\,mJy\,$\nu ^{1.33}$

\begin{figure}
\includegraphics[trim = 0mm 0mm 0mm 5mm, clip,scale=0.38, angle=90]{/home/eamon/pietown/paper/figs/fig3.ps}
\caption{xxx }
\label{fig3}
\end{figure}

\subsection{Temperature Profiles}
give more exact lim formula

@@

\subsection{Structure of Wind Acceleration Region} 
Thermal structure (Lim vs vs Ours vs e-MERLIN)\\
see Reid \& Menton 1996 conf proceedings\\
Harper model\\
Highlighted skinner discussion on velocities needed to produce a change in size
